---
title: "Analysis #2: District policy and health department-reported case data"
format: 
  html:
    page-layout: full
    self-contained: true  
    mainfont: "Times New Roman"
    theme: journal
editor: visual
---

```{r}
#| echo: false
#| include: false
library(DT)
library(dplyr)
library(tidyr)
readRDS("raw_summary_statistics.rds")%>%
  list2env(envir=.GlobalEnv)

readRDS("output_tables.rds")%>%
  list2env(envir=.GlobalEnv)
```

# Introduction

The purpose of this analysis is to study the effectiveness of school districts' COVID-19 prevention strategies in mitigating the spread of COVID-19 among school-aged children and adolescents. Specifically, for the 2021 - 2022 school year, we collected 1,184 COVID-19 policy documents from the websites of a nationally-representative random sample of 1,286 school districts, as well as monthly case counts from state health departments. As such, the primary research question is, what is the association between districts' COVID-19 prevention policies and subsequent COVID-19 caseloads in schools?

# Methods

**Outcome of interest**: Difference between schools' fall and spring monthly average case counts, represented as the difference between the average of schools' monthly number of cases across October, November, and December 2021 and average number of monthly cases for January, February, and March 2022.

**Independent variables:** The key predictors of interest consist of 10 categories of COVID-19 mitigation strategies, including:

-   *Vaccination*: Establish policies and practices that promote vaccinations among teachers, staff, families, and eligible students by providing information about COVID-19, encouraging vaccine trust and confidence, and making getting vaccinated as easy and convenient as possible. 
-   *Handwashing and Resporitory Etiquette*: Practice handwashing and respiratory etiquette (covering coughs and sneezes) to keep from getting and spreading COVID-19. Teach and reinforce handwashing with soap and water for at least 20 seconds. Remind everyone in the facility to wash hands frequently, and adults assist young children with handwashing. If handwashing is not possible, use hand sanitizer containing at least 60% alcohol. 
-   *Masking*: Teachers, staff, students, and visitors wear masks consistently and correctly (i.e., covering the mouth and nose) indoors regardless of vaccination status for all individuals ages 2 years and older. Unvaccinated people also wear masks outdoors in crowded spaces. All individuals, including students and drivers, wear masks on school buses. 
-   *Physical distancing*: Schools maintain at least 3 feet of physical distance between students within classrooms. A distance of at least 6 feet is recommended between students and teachers/staff and between teachers/staff who are not fully vaccinated. 
-   *Cohorting/Staggering*: Keep people together in a small group and have each group stay together throughout an entire day to limit the number of students, teachers, and staff who come in contact with each other. 
-   *Testing and screening*: Screen and test to identify infected people, including those with or without symptoms (or before development of symptoms) who may be contagious, so that measures can be taken to prevent further transmission. At a minimum, screening testing should be offered to students who have not been fully vaccinated when community transmission is at moderate, substantial, or high levels. At all levels of community transmission, screening testing should be offered to all teachers and staff who have not been fully vaccinated.  
-   *Staying home*: Anyone with COVID-19 stays home and isolated from other people for 10 days. Schools should provide excused absences for students who are sick. Schools should also allow flexible, non-punitive, and supportive paid sick leave policies and practices that encourage sick workers to stay home without fear of retaliation, loss of pay, or loss of employment level.  
-   *Trace and quarantine*: Collaborate with state and local health departments to the extent allowable by privacy and other applicable laws to confidentially provide information about people diagnosed with or exposed to COVID-19. While students, teachers, and staff with positive COVID-19 test results should isolate, contact tracing helps identify close contacts who should quarantine and closely monitor for symptoms.
-   *Cleaning*: Clean school surfaces at least once a day to remove potential virus particles. Use disinfectants on the U.S. Environmental Protection Agency COVID-19 list to kill any remaining germs on surfaces. 
-   *Ventilation*: Bring fresh outdoor air into a building to help keep virus particles from concentrating inside. This can be done by opening doors and windows, using child-safe fans to increase circulation, and making changes to HVAC or air filtration systems. Open or crack windows in buses and other forms of transportation if doing so does not pose a safety risk. 

*School-level characteristics*

-   *Percent student body eligible for free and reduced lunch*
-   *School locale*: City, Rural, Suburb, Town, Missing
-   *Percent of student body Asian, American Indian or Alaska Native, Black/African American, Hispanic/Latino, Native Hawaiian or other Pacific Islander, Not specific, Two or more races, and White (each race/ethnicity represented individually)*

*County-level characteristics*

-   *Social Vulnerability Index (SVI)*: Overall summary index indicating the relative vulnerability of U.S. Census tracts across four themes: socioeconomic, household composition & disability, minority status & language, and housing type & transportation
-   *Change in county COVID-19 case rates*: calculated as difference in average monthly case rate per 100,000 people between October - December and January - March, corresponding with the time period used for calculating school case rate changes
-   *Region:* Midwest, Northeast, South, West
-   *State*

# Results

As we explore the data there are multiple avenues to framing district policies, one prominent characteristic being *when* policies were enacted. For the current study, we distinguish between policies enacted during the fall and spring semesters, re-scoring those enacted during the spring semester as not being in place. Given that the policies were scored on continuous scales (see table 2), they can also be assessed as categorical (i.e., dichotomous) indicators. For this initial review, policies are dichotimized by zero/nonzero scores and whether policy scored in 75th percentile, yielding three sets of summaries: continuous, comparison of no policy/policy, and policies below and at or above 75th percentile. For continuous variables, Pearson's correlation coefficient is calculated as an initial measure of association with the change in case rates between spring and fall (see tables 1 and 2). For dichotomous indicators, t-tests are conducted, comparing the average change in case rates between schools with and without the strategies in place (table 3 for no policy/policy; table 4 for 75th percentile).

More specifically, initial results are summarized as follows. First, table 1 includes summary statistics of the outcome of interest (change in case rate) as well as covariates. Table 2 provides the summary statistics, including Pearson's correlation coefficients, for strategies operationalized as a continuous variable (0 meaning no strategy in place; max indicating the max score possible for a given strategy). Tables 3 and 4 review summary statistics and t-test results from strategies dichotimized as 0 if not having a policy or enactment during spring/below 75th percentiles, and 1 if having any policy during the fall/scoring at 75th percentile or greater.

```{r}
#| echo: false
tbl1 %>%
   DT::datatable(rownames = FALSE,colnames=c("n (min, max)"='min_max',"Mean (median)"="mean_median", "Correlation (p-value)"="est_pval"),
                 style="bootstrap4",
                 caption = "Table 1: Summary statistics of changes in case rates and standardized covariates",
                extensions = 'Buttons', options = list(
                  pageLength=nrow(tbl1),
    dom = 'tB',
    buttons = c('excel', 'copy'))
                 )
```

```{r}
#| echo: false
tbl2 %>%
   DT::datatable(rownames = FALSE,colnames=c("n (min, max)"='min_max',"Mean (median)"="mean_median", "Correlation (p-value)"="est_pval"),
                 style="bootstrap4",
                 caption = "Table 2: Summary statistics of strategies treated as continuous and correlations with changes in case rates",
                extensions = 'Buttons', options = list(
    pageLength=nrow(tbl2),
    dom = 'tB',
    buttons = c('excel', 'copy'))
                 )
```

```{r}
#| echo: false
tbl3 %>%
   DT::datatable(rownames = FALSE,colnames=c("n (min, max)"='min_max',"Mean (ST Dev)"="mean_stdev", "Difference in means (p-value)"="t_estimate", 'No policy'='No policy_mean','Has policy'='Has policy_mean'),
                 style="bootstrap4",
                 caption = "Table 3: Summary statistics and t-test results of dichotomized policies: Comparing no-policy to policy",
                extensions = 'Buttons', options = list(
    pageLength=nrow(tbl3),
    dom = 't B',
    buttons = c('excel', 'copy'))
                 )
```

```{r}
#| echo: false
tbl4 %>%
   DT::datatable(rownames = FALSE,colnames=c("n (min, max)"='min_max',"Mean (ST Dev)"="mean_stdev", "Difference in means (p-value)"="t_estimate", 'No policy'='No policy_mean','Has policy'='Has policy_mean'),
                 style="bootstrap4",
                 caption = "Table 4: Summary statistics and t-test results of policies dichotimized by scoring at or above 75th percentile",
                extensions = 'Buttons', options = list(
    pageLength=nrow(tbl3),
    dom = 'tB',
    buttons = c('excel', 'copy'))
                 )
```

# Discussion
